Revisiting Child and Adolescent Depression Treatments

### 2024-05-13 Manuscript_for_rendering_in_word.qmd
Removed TADS from baseline comparisons Calculated between group effect sizes for psy active vs control

We included RCTs identified in a recent meta-analysis of psychotherapy versus control [@cuijpersEffectsPsychologicalTreatments2021] and an NMA examining the efficacy of antidepressants [@ciprianiComparativeEfficacyTolerability2016] for depression in children and adolescents. For the psychotherapy trials, we utilised open data from the previous meta-analysis (see [here](https://www.metapsy.org/)). For medication trials, we were unable to access the full dataset used in the NMA and hence extracted data from the included studies ourselves. We excluded three studies because they had no control arm. We were unable to locate and complete extraction for two RCTs (Almeida-Montes, 2005; Eli Lilly, 1986). Many studies did not report complete data; we contacted all corresponding authors to request missing data, though did not receive any responses.

@@

As the measure of effect of each individual study, we used the within-group Standardised Mean Difference (SMD).

@@

df_for_baseline_severity <- df_for_baseline_severity %>% 
  distinct(new_study_id, .keep_all = TRUE)

@@

df_for_sex_metan <- df_for_sex_metan %>% 
  filter(percent_women != 100)

@@

df_for_age_metan <- df_baseline_demographics %>% 
  filter(!is.na(mean_age))

weighted_average_age <- sum(df_baseline_demographics$mean_age * df_baseline_demographics$baseline_n, na.rm = TRUE) / sum(df_baseline_demographics$baseline_n, na.rm = TRUE)
tst <- mean(df_baseline_demographics$mean_age, na.rm = TRUE)

@@

```
We addressed potential regression to the mean by residualising SMDs by the percentage adjusted baseline scores in one regression model (see @tbl-adj-smds in the Supplementary Materials for the adjusted SMDs with their confidence intervals). The difference between the medication control and psychotherapy control arms remained similar to the overall analyses.

@@

df_baseline_demographics <- full_join(df_baseline_demographics, df_no_sites, by = "new_study_id")

@@

sites_t_test_stats <- sites_t_test_stats %>%
  mutate(across(.cols = c(statistic, df, ci_lower, ci_upper), ~ round(., 2))) %>% 
  mutate(p_value = ifelse(p_value < 0.001, "< 0.001", as.character(round(p_value, 3))))
```

@@

As shown in Average number of trial sites was significantly higher in medication trials (*M* = `r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Medication"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Medication"]`) compared to psychotherapy studies (*M* =`r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Psychotherapy"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Psychotherapy"]`)(*t* (`r round (sites_t_test$parameter, 2)`) = `r round(sites_t_test$statistic,2)`, *p* =`r sites_t_test_stats$p_value`). Of those studies for which we had data on number of sites, 26 of 28 (93%) medication trials were multisite, compared to 24 of 45 (54%) psychotherapy studies.

@@

control_summary <- control_summary %>% 
  mutate(across(where(is.numeric), ~round(., 2)))
```

@@

@@

We drew upon RCTs included in two recent comprehensive meta-analyses with open data available for each medication and psychotherapy, and supplemented them with an updated systematic review. Please refer to these original meta-analyses for a detailed description of their search strategy and study selection criteria. Psychotherapy studies were drawn from a systematic review and meta-analysis of randomised trials comparing psychotherapy for youth depression against control conditions [@cuijpersEffectsPsychologicalTreatments2021] (dataset available [here](https://www.metapsy.org/)). Whilst Cuijpers et al. (2021) excluded studies for which the primary outcome variable could not be calculated due to missing data, we included these studies and performed the imputations outlined below. We also included studies which had data available for other variables in interest, including number of sites or baseline demographics; hence we have more psychotherapy studies included in this review compared to the original meta-analysis. Whilst the online database is regularly updated, we chose to exclude studies published after the final date of Cuijpers et al.'s (2021) literature search.

@@

control_summary_no_wl <- control_summary_no_wl %>% 
  mutate(across(where(is.numeric), ~round(., 2)))
```

>>>>

We included RCTs identified in a recent meta-analysis of psychotherapy versus control [@cuijpersEffectsPsychologicalTreatments2021] and an NMA examining the efficacy of antidepressants [@ciprianiComparativeEfficacyTolerability2016] for depression in children and adolescents. For the psychotherapy trials, we utilised open data from the previous meta-analysis (see [here](https://docs.metapsy.org/databases/depression-childadol-psyctr/)). For medication trials, we were unable to access the full dataset used in the NMA and hence extracted data from the included studies ourselves. We excluded three studies because they had no control arm. We were unable to locate and complete extraction for two RCTs (Almeida-Montes, 2005; Eli Lilly, 1986). Many studies did not report complete data; we contacted all corresponding authors to request missing data, though did not receive any responses.

@@

As the measure of effect of each individual study, we used the within-group Standardised Mean Difference (SMD) for the primary depression symptom scale used (selected according to the hierarchy defined in the Supplement).

@@

df_for_baseline_severity <- df_for_baseline_severity %>% 
  distinct(new_study_id, .keep_all = TRUE) %>% 
  filter(new_study_id != "March, 2004") # removing TADS because it does not technically belong to either the psy or med groups

@@

df_for_sex_metan <- df_for_sex_metan %>% 
  filter(percent_women != 100) # filter out studies with only women %>% 
  filter(new_study_id != "March, 2004") # removing TADS because it does not technically belong to either the psy or med groups

@@

df_for_age_metan <- df_baseline_demographics %>% 
  filter(!is.na(mean_age)) %>% 
  filter(new_study_id != "March, 2004") # removing TADS because it does not technically belong to either the psy or med groups

@@

We addressed potential regression to the mean by residualising SMDs by the percentage adjusted baseline scores in one regression model (see @tbl-adj-smds in the Supplementary Materials for the adjusted SMDs with their confidence intervals). The difference between the medication control and psychotherapy control arms remained similar to the overall analyses.

@@

df_baseline_demographics <- full_join(df_baseline_demographics, df_no_sites, by = "new_study_id") %>% 
  filter(new_study_id != "March, 2004") # removing TADS because it does not technically belong to either the psy or med groups

@@

sites_t_test_stats <- sites_t_test_stats %>%
  mutate(across(.cols = c(statistic, df, ci_lower, ci_upper), ~ round(., 2))) %>% 
  mutate(p_value = ifelse(p_value < 0.001, "< 0.001", as.character(round(p_value, 3))))

@@

Average number of trial sites was significantly higher in medication trials (*M* = `r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Medication"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Medication"]`) compared to psychotherapy studies (*M* =`r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Psychotherapy"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Psychotherapy"]`)(*t* (`r round (sites_t_test$parameter, 2)`) = `r round(sites_t_test$statistic,2)`, *p* =`r sites_t_test_stats$p_value`). Of those studies for which we had data on number of sites, 26 of 28 (93%) medication trials were multisite, compared to 24 of 45 (54%) psychotherapy studies.

@@

control_summary <- control_summary %>% 
  mutate(across(where(is.numeric), ~round(., 2)))

#function for between group effect sizes (cohens d)

effect_btwn <- function(df) {
  pooled_sd <- sqrt(((df$N[1]-1)*df$SD[1]^2 + (df$N[2]-1)*df$SD[2]^2) / (df$N[1] + df$N[2] - 2))
  mean_diff <- df$Mean[1] - df$Mean[2]
  effect_btwn <- mean_diff / pooled_sd
  return(effect_btwn)
}

session_es <- effect_btwn(sessions)
frequency_es <- effect_btwn(frequency)
length_sessions_mins_es <- effect_btwn(length_sessions_mins)
total_hours_es <- effect_btwn(total_hours)
total_period_weeks_es <- effect_btwn(total_period_weeks)

view(df_descr_psy_conditions)
testing <- df_descr_psy_conditions %>% 
  filter()

other_ctr_ids <- df_long_for_metan[df_long_for_metan$type == "other ctr",]$new_study_id
testing <- df_descr_psy_conditions %>% 
  filter(new_study_id %in% other_ctr_ids)
view(testing)
```

@@

**References for included trials**

References for trials included in the existing meta-analyses drawn upon for this study can be found [here](https://docs.metapsy.org/databases/depression-childadol-psyctr/) for psychotherapy RCTs and [here](https://ora.ox.ac.uk/objects/uuid:e0b5ae23-d562-4348-94b8-84f70b7812c5) for medication RCTs. Below are references for the seven additional RCTs identified in the original systematic review conducted for the current study.

*Publications*

Atkinson, S., Lubaczewski, S., Ramaker, S., England, R. D., Wajsbrot, D. B., Abbas, R., & Findling, R. L. (2018). Desvenlafaxine versus placebo in the treatment of children and adolescents with major depressive disorder. *Journal of Child and Adolescent Psychopharmacology, 28(1),* 55-65. DOI: 10.1089/cap.2017.0099

Durgam, S., Chen, C. Z., Migliore, R., Prakash, C., Edwards, J., & Findling, R. L. (2018). A phase 3, double-blind, randomized, placebo-controlled study of vilazodone in adolescents with major depressive disorder. *Pediatric Drugs, 20(4),* 353-363. DOI: 10.1007/s40272-018-0290-4

Findling, R. L., McCusker, E., & Strawn, J. R. (2020). A randomized, double-blind, placebo-controlled trial of vilazodone in children and adolescents with major depressive disorder with twenty-six-week open-label follow-up. *Journal of Child and Adolescent Psychopharmacology, 30(6),* 355-365. DOI: 10.1089/cap.2019.0176

Le Noury, J., Nardo, J. M., Healy, D., Jureidini, J., Raven, M., Tufanaru, C., & Abi-Jaoude, E. (2015). Restoring Study 329: efficacy and harms of paroxetine and imipramine in treatment of major depression in adolescence. *BMJ, 351,*, h4320. DOI: 10.1136/bmj.h4320

Weihs, K. L., Murphy, W., Abbas, R., Chiles, D., England, R. D., Ramaker, S., & Wajsbrot, D. B. (2018). Desvenlafaxine versus placebo in a fluoxetine-referenced study of children and adolescents with major depressive disorder. Journal of Child and Adolescent Psychopharmacology, 28(1), 36-46. DOI: 10.1089/cap.2017.0100

*Unpublished clinical trials*

Active Reference (Fluoxetine) Fixed-dose Study of Vortioxetine in Paediatric Patients Aged 12 to 17 Years With Major Depressive Disorder (MDD). ClinicalTrials.gov Identifier: NCT02709746. Accessed 17 Jan 2024.

Safety and Efficacy of Levomilnacipran ER in Adolescent Participants With Major Depressive Disorder. ClinicalTrials.gov Identifier: NCT02431806. Accessed 17 Jan 2024.

**Hierarchy of depression symptom severity measurement scales**

Where multiple depression rating scales were used, we selected the best available measure according to the following hierarchy used in Cipriani et al. (2016).

1. Children’s Depression Rating Scale (CDRS)
1. Hamilton Depression Rating Scale (HAMD)
1. Montgomery Asberg Depression Rating Scale (MADRS)
1. Beck Depression Inventory (BDI)
1. Children’s Depression Inventory (CDI)
1. Schedule for Affective Disorders and Schizophrenia for School-Aged Children (K-SADS)
1. Mood and Feeling Questionnaire (MFQ)
1. Reynolds Adolescent Depression Scale (RADS)
1. Bellevue Index of Depression (BID)
1. Child Depression Scale (CDS)
1. Centre for Epidemiological Studies Depression Scale (CES-D)
1. Child Assessment Schedule (CAS)
1. Child Behaviour Checklist-Depression (CBCL-D)

@@

We drew upon RCTs included in two recent comprehensive meta-analyses with open data available for each medication and psychotherapy, and supplemented them with an updated systematic review. Please refer to these original meta-analyses for a detailed description of their search strategy and study selection criteria. Psychotherapy studies were drawn from a systematic review and meta-analysis of randomised trials comparing psychotherapy for youth depression against control conditions [@cuijpersEffectsPsychologicalTreatments2021] (dataset available [here](https://docs.metapsy.org/databases/depression-childadol-psyctr/)). Whilst Cuijpers et al. (2021) excluded studies for which the primary outcome variable could not be calculated due to missing data, we included these studies and performed the imputations outlined below. We also included studies which had data available for other variables in interest, including number of sites or baseline demographics; hence we have more psychotherapy studies included in this review compared to the original meta-analysis. Whilst the online database is regularly updated, we chose to exclude studies published after the final date of Cuijpers et al.'s (2021) literature search.

@@

control_summary_no_wl <- control_summary_no_wl %>% 
  mutate(across(where(is.numeric), ~round(., 2)))

session_es_no_wl <- effect_btwn(sessions)
frequency_es_no_wl <- effect_btwn(frequency)
length_sessions_mins_es_no_wl <- effect_btwn(length_sessions_mins)
total_hours_es_no_wl <- effect_btwn(total_hours)
total_period_weeks_es_no_wl <- effect_btwn(total_period_weeks)
